### 2017-05-06 2017mergingmerges/section/02-Processes.tex
Fix sigma -> bs in advance rules

\subsection{Breaking It Down}
Overall, we started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph. Importantly, we needed to ``break down'' the definition of each of the operators into imperative statements so that we could interleave their execution appropriately. We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way. Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@. Instead, in the fused result program we have effectively performed the job of a concurrent scheduler at compile time. The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.

The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative code that implements the operators was also used in prior work on Flow Fusion \cite{lippmeier2013data}. However, in contrast to Flow Fusion and similar systems, with Machine Fusion we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss in \S\ref{s:FutureWork}.

>>>>

\subsection{Breaking It Down}
We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph. Importantly, we needed to \emph{break down} the definition of each of the operators into imperative statements so that we could interleave their execution appropriately. We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way. Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@. Instead, in the fused result program we have performed the job of a concurrent scheduler at compile time. The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.

The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative code that implements the operators was also used in prior work on Flow Fusion~\cite{lippmeier2013data}. However, in contrast to Flow Fusion and similar systems, with \mbox{Machine Fusion} we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss in \S\ref{s:FutureWork}.

### 2017-05-06 2017mergingmerges/section/03-Definition.tex
Fix sigma -> bs in advance rules

The grammar for process definitions is given in Fig.~\ref{fig:Process:Def}. Variables, Channels and Labels are specified by unique names. We refer to the \emph{endpoint} of a stream as a channel. A particular stream may flow into the input channels of several different processes, but can only be produced by a single output channel. For values and expressions we use an untyped lambda calculus with a few primitives chosen to facilitate the examples. The `$||$' operator is boolean-or, `+' addition, `/=' not-equal, and `$<$' less-then.

A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the instruction currently being executed, and the @instrs@ a map of labels to instructions. We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics. When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed. Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$. If an $\InputState$ is not specified we assume it is `none'.
A network is a set of processes, able to communicate to each other.

In the grammar of Fig.~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable. The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes. When performing the fusion transform the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state. When defining the fusion transform in \S\ref{s:Fusion} we will used a version of $\InputState$ with only this staticaly known information.

@@

When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely convert @drop@ to plain @jump@ instructions. The @drop@s instructions are used to control how processes should be synchronized, but do not affect the execution of a single process. We will discuss @drop@s further in \S\ref{s:Optimisation}.

@@

Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not. When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the network to advance without requiring unbounded buffering.

@@

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, operating on a network.

@@

% -----------------------------------------------------------------------------
\subsubsection{Advancing}
Fig.~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process. The first set of rules handle specific instructions. The statement $(\ProcBlockShake{i}{is}{bs}{a}{l}{is'}{us'})$ reads ``instruction $i$, given channel states $is$ and the heap bindings $bs$, passes control to instruction at label $l$ and yields new channel states $is'$, heap update expressions $us'$, and performs an output action $a$.'' An output action $a$ is an optional message of the form $(\Push~\Chan~\Value)$, which encodes the value a process pushes to one of its output channels. We write ~$\cdot$~ for an empty action.

@@

% -----------------------------------------------------------------------------
\subsubsection{Feeding}
Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes. The first set of rules concerns feeding values to other processes within the same network, while the second exchanges input and output values with the environment the network is running in.

@@

The execution rules of Fig. \ref{fig:Process:Eval:Feed} are non-deterministic in several ways. Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the network; (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the network are ready to accept the pushed value; (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the channel are ready to accept the value.

>>>>

The formal grammar for process definitions is given in Figure~\ref{fig:Process:Def}. Variables, Channels and Labels are specified by unique names. We refer to the \emph{endpoint} of a stream as a channel. A particular stream may flow into the input channels of several different processes, but can only be produced by a single output channel. For values and expressions we use an untyped lambda calculus with a few primitives chosen to facilitate the examples. The `$||$' operator is boolean-or, `+' addition, `/=' not-equal, and `$<$' less-then.

A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the instruction currently being executed, and the @instrs@ a map of labels to instructions. We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics. When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed. Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$. If an $\InputState$ is not specified we assume it is `none'. A network is a set of processes that are able to communicate with each other.

In the grammar of Figure~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable. The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes. When performing the fusion transform the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state. When defining the fusion transform in \S\ref{s:Fusion} we will use a version of $\InputState$ with only this staticaly known information.

@@

When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely convert @drop@ to plain @jump@ instructions. The @drop@ instructions are used to control how processes should be synchronized, but do not affect the execution of a single process. We discuss @drop@s further in \S\ref{s:Optimisation}.

@@

% TODO BL: Mention what happens if we choose a different ordering, 
% and how the particular ordering chosen is decided upon.
Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not. When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the network to advance without requiring unbounded buffering.

@@

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Figure~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, operating on a network.

@@

% -----------------------------------------------------------------------------
\subsubsection{Advancing}
Figure~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process. The first set of rules handle specific instructions. The statement $(\ProcBlockShake{i}{is}{bs}{a}{l}{is'}{us'})$ reads ``instruction $i$, given channel states $is$ and the heap bindings $bs$, passes control to instruction at label $l$ and yields new channel states $is'$, heap update expressions $us'$, and performs an output action $a$.'' An output action $a$ is an optional message of the form $(\Push~\Chan~\Value)$, which encodes the value a process pushes to one of its output channels. We write ~$\cdot$~ for an empty action.

@@

% -----------------------------------------------------------------------------
\subsubsection{Feeding}
Figure~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes. The first set of rules concerns feeding values to other processes within the same network, while the second exchanges input and output values with the environment the network is running in.

@@

The execution rules of Figure~\ref{fig:Process:Eval:Feed} are non-deterministic in several ways. Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the network; (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the network are ready to accept the pushed value; (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the channel are ready to accept the value.
